---
title: "Functions of covariates and evaluation environments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions of covariates and evaluation environments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gratia)
library(mgcv)
```

A smooth can depend on a function of a covariate rather than on the raw
covariate. The same distinction arises with offsets, such as logged effort
in a count model.

```{r model}
set.seed(42)
dat <- data.frame(x = runif(150, 1, 5), effort = runif(150, 1, 3))
dat$y <- rpois(nrow(dat), dat$effort * exp(sin(log(dat$x))))
m <- gam(y ~ s(log(x), k = 6) + offset(log(effort)),
  family = poisson(), data = dat, method = "REML")
```

## Raw data and smooth coordinates

Supply raw covariates to evaluate the model at new observations. Gratia
computes `log(x)` and, for full predictions, `offset(log(effort))`. An isolated
smooth does not require unrelated predictors or the model offset.

```{r raw-data}
nd <- data.frame(x = seq(1.2, 4.8, length.out = 8), effort = 2)
smooth_estimates(m, data = nd["x"])
fitted_values(m, data = nd)
```

Automatic smooth grids are evenly spaced in the **smooth coordinate**, here
`log(x)`, and the output and plot axes use that name. They use stored evaluated
covariates, so they do not need to reconstruct an inverse transformation.

```{r coordinates}
sm <- smooth_estimates(m, n = 30)
draw(sm)
```

An explicitly evaluated column is also accepted:

```{r evaluated-data}
evaluated <- data.frame(`log(x)` = log(nd$x), check.names = FALSE)
smooth_estimates(m, data = evaluated)
```

If raw inputs and evaluated columns are both supplied, expressions are
recomputed from the raw inputs. Stored model frames and automatically prepared
grids retain their evaluated coordinates.

`data_slice()` works with raw variables. Supply reference `data` if the original
raw covariates cannot be recovered; transformed values are not renamed as raw
variables or automatically inverted.

```{r raw-slice}
data_slice(m, x = evenly(x, n = 8), effort = 2, data = dat)
```

## Local functions

A fitted model may retain the formula expression without retaining the local
environment in which it was evaluated. Preserve that environment when fitting
if you will need a local function at new observations.

```{r local-function}
fit_local <- function(dat) {
  shift <- 0.5
  transform_x <- function(x) log(x + shift)
  f <- y ~ s(transform_x(x), k = 6) + offset(log(effort))
  list(model = gam(f, family = poisson(), data = dat, method = "REML"),
    envir = environment(f))
}
bundle <- fit_local(dat)
smooth_estimates(bundle$model, data = nd, envir = bundle$envir)
fitted_values(bundle$model, data = nd, envir = bundle$envir)
```

Automatic smooth plots can still use stored `transform_x(x)` values without
this environment. New raw observations require the function. Evaluation errors
identify the expression and explain how to supply the missing context; stored
training values are never substituted for arbitrary new observations.

## Which derivative?

For a contribution $f(\log x)$, the default derivative is with respect to
$\log x$. Set `wrt = "covariate"` to differentiate with respect to $x$ instead.
The latter includes the chain-rule factor $1/x$ because the transformation is
reevaluated after each perturbation of $x$.

```{r derivatives}
derivatives(m, data = nd, type = "central")
derivatives(m, data = nd, type = "central", wrt = "covariate")
```

For a smooth of several raw inputs, such as `s(I(x / z))`, name the raw
`focal` variable explicitly. Other inputs are held at the supplied values, or
at typical values when generating a slice. Use `I()` for arithmetic that R
would otherwise interpret as formula syntax. For tensor smooths, use
`partial_derivatives()` and specify the smooth or raw focal coordinate
consistently with `wrt`.
