---
title: "Functions of covariates and evaluation environments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions of covariates and evaluation environments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gratia)
library(mgcv)
```

A smooth can depend on a function of a covariate rather than on the raw
covariate. The same distinction arises with offsets, such as logged effort
in a count model.

```{r model}
set.seed(42)
dat <- data.frame(x = runif(150, 1, 5), effort = runif(150, 1, 3))
dat$y <- rpois(nrow(dat), dat$effort * exp(sin(log(dat$x))))
m <- gam(y ~ s(log(x), k = 6) + offset(log(effort)),
  family = poisson(), data = dat, method = "REML")
```

## Raw data and smooth coordinates

Supply raw covariates to evaluate the model at new observations. Gratia
computes `log(x)` and, for full predictions, `offset(log(effort))`. An isolated
smooth does not require unrelated predictors or the model offset.

```{r raw-data}
nd <- data.frame(x = seq(1.2, 4.8, length.out = 8), effort = 2)
smooth_estimates(m, data = nd["x"])
fitted_values(m, data = nd)
```

Automatic smooth grids are evenly spaced in the **smooth coordinate**, here
`log(x)`, and the output and plot axes use that name. They use stored evaluated
covariates, so they do not need to reconstruct an inverse transformation.

```{r coordinates}
sm <- smooth_estimates(m, n = 30)
draw(sm)
```

An explicitly evaluated column is also accepted:

```{r evaluated-data}
evaluated <- data.frame(`log(x)` = log(nd$x), check.names = FALSE)
smooth_estimates(m, data = evaluated)
```

If raw inputs and evaluated columns are both supplied, expressions are
recomputed from the raw inputs. Stored model frames and automatically prepared
grids retain their evaluated coordinates.

`data_slice()` works with raw variables. Supply reference `data` if the original
raw covariates cannot be recovered; transformed values are not renamed as raw
variables or automatically inverted.

```{r raw-slice}
data_slice(m, x = evenly(x, n = 8), effort = 2, data = dat)
```

## Local functions

A fitted model may retain the formula expression without retaining the local
environment in which it was evaluated. Preserve that environment when fitting
if you will need a local function at new observations.

```{r local-function}
fit_local <- function(dat) {
  shift <- 0.5
  transform_x <- function(x) log(x + shift)
  f <- y ~ s(transform_x(x), k = 6) + offset(log(effort))
  list(model = gam(f, family = poisson(), data = dat, method = "REML"),
    envir = environment(f))
}
bundle <- fit_local(dat)
smooth_estimates(bundle$model, data = nd, envir = bundle$envir)
fitted_values(bundle$model, data = nd, envir = bundle$envir)
```

Automatic smooth plots can still use stored `transform_x(x)` values without
this environment. New raw observations require the function. Evaluation errors
identify the expression and explain how to supply the missing context; stored
training values are never substituted for arbitrary new observations.

## Which derivative?

For a contribution $f(\log x)$, the default derivative is with respect to
$\log x$. Set `wrt = "covariate"` to differentiate with respect to $x$ instead.
The latter includes the chain-rule factor $1/x$ because the transformation is
reevaluated after each perturbation of $x$.

```{r derivatives}
derivatives(m, data = nd, type = "central")
derivatives(m, data = nd, type = "central", wrt = "covariate")
```

For a smooth of several raw inputs, such as `s(I(x / z))`, name the raw
`focal` variable explicitly. Other inputs are held at the supplied values, or
at typical values when generating a slice. Use `I()` for arithmetic that R
would otherwise interpret as formula syntax. For tensor smooths, use
`partial_derivatives()` and specify the smooth or raw focal coordinate
consistently with `wrt`: **`wrt` chooses the differentiation scale, and `focal`
names the coordinate to differentiate along on that scale**.

For example, `te(log(x), sqrt(z))` represents a contribution
$f(\log x, \sqrt z)$. The two choices have the following meanings:

| Arguments | Derivative |
|:----------|:-----------|
| `wrt = "smooth", focal = "log(x)"` | With respect to $\log x$, holding $\sqrt z$ fixed |
| `wrt = "covariate", focal = "x"` | With respect to $x$, holding $z$ fixed |

```{r partial-derivatives}
set.seed(43)
tensor_dat <- data.frame(x = runif(150, 1, 5), z = runif(150, 1, 3))
tensor_dat$y <- sin(log(tensor_dat$x)) * sqrt(tensor_dat$z) +
  rnorm(nrow(tensor_dat), sd = 0.1)
tensor_model <- gam(y ~ te(log(x), sqrt(z), k = c(4, 4)),
  data = tensor_dat, method = "REML")

# Vary x along a slice with z held fixed.
tensor_nd <- data.frame(x = seq(1.2, 4.8, length.out = 8), z = 2)
pd_smooth <- partial_derivatives(tensor_model, data = tensor_nd,
  wrt = "smooth", focal = "log(x)", type = "central")
pd_covariate <- partial_derivatives(tensor_model, data = tensor_nd,
  wrt = "covariate", focal = "x", type = "central")

# Compare the raw-covariate derivative with the chain-rule calculation.
data.frame(x = tensor_nd$x,
  covariate_derivative = pd_covariate$.partial_deriv,
  smooth_derivative_over_x = pd_smooth$.partial_deriv / tensor_nd$x)
```

The first call perturbs `log(x)` directly. The second perturbs `x` and
recomputes the smooth expressions, incorporating the chain rule numerically.
Thus, for these first derivatives,
$\partial f / \partial x = (\partial f / \partial \log x) / x$,
up to finite-difference approximation error.

In `partial_derivatives()`, `wrt = "smooth"` is the default, and omitting
`focal` selects the first non-factor smooth coordinate. With
`wrt = "covariate"`, supply `focal` whenever several raw covariates are
eligible; for this tensor smooth, choose `"x"` or `"z"`. If selecting several
smooths, supply one `focal` per smooth in their selected order; the same `wrt`
applies to all of them.

Supplied `data` should describe a slice with the other independent numeric
inputs held fixed, as `z` is above. Raw-scale differentiation requires the raw
focal column: stored transformed values alone are insufficient. For an
untransformed coordinate such as `x` in `te(x, z)`, both choices of `wrt` give
the same interpretation.
